---
title: "Making Graphs in R Using ggplot2: Part 1"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    df_print: default
runtime: shiny_prerendered
description: >
  Learn about the basic syntax of R.
---

## Overview

![](https://youtu.be/lbt4BH9Q82E)

In this session we will start looking at one of the most widely used, and almost universally useful, applications of the R programming languages - data visualisation. We will do this by learning specifically about the plotting library `ggplot2`. There are three parts to this workbook, as well as a series of exercises and questions. You can complete the workbooks at your own pace, across multiple sessions, although note that progress through each part of the workbook only saves if you are using the same device. However you can skip back and forth at any point within each part using the navigation menu.

### Installing the ggplot2 package
As mentioned in the video, this is not the default plotting library within R - but an additional package which needs to be installed separately. We will learn more about installing packages a bit later in the course, but an important thing to remember is that when you are using RStudio or R, to make sure you have downloaded the package first,using `install.packages("ggplot2")`, and that you have loaded the package for the session, using `library(ggplot2)`. The package has already been pre-loaded into this workbook.

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(knitr)
tutorial_options(exercise.timelimit = 10)
BeanSurvey<-read.csv("bean_survey.csv")[,-1]

BeanSurvey$AGEHH<-c(32,57,20,55,78,42,28,49,41,54,65,63,32,29,43,68,53,39,25,44,23,47,32,26,25,29,44,23,58,69,26,65,75,51,38,24,35,29,37,45,50,23,60,32,70,43,51,28,65,33)


BeanSurvey<-mutate(BeanSurvey,GENDERHH=ifelse(is.na(GENDERHH),"male",as.character(GENDERHH)))


```



## 'BeanSurvey' dataset 

The data we are using in this session is an extract of a survey conducted in Uganda from farmers identified as growing beans.

The dataset contains an extract of 50 responses to 23 of the survey questions, and has been imported to R as a data frame called `BeanSurvey`.

A summary of the columns in the dataset is below.

```{r data1, echo=FALSE,message=FALSE,warning=FALSE}


data.frame(Column=colnames(BeanSurvey),
           Description=c("Farmer ID","Village name","Household composition","Gender of Household Head","Age of Household Head",
                         "Occupation of Household Head","Number of Adults within the household","Number of Children (<18) within the household","Do they grow matoke?",
                         "Do they grow maize?","Do they grow beans?","Do they grow banana?","Do they grow cassava?",
                         "Do they grow coffee?","Land area of farm (acres)","Labor usage","Intercrops with beans",
                         "Household decision responsibility","Do they grow beans for sale?","Quantity of beans planted in long rain season",
                         "Quantity of beans planted in short rain season","Quantity of beans harvested in long rain season"
                         ,"Quantity of beans harvested in short rain season")) %>% kable()

```


Spend some time looking through the exploring the full dataset embedded below, to familiarise yourself with the columns and the type of data stored within each column. You may need to refer back to this data at times during this tutorial. Remember that R is case sensitive, so you will always have to refer to the variables in this dataset exactly as they are written in the data. There is a column in this data called "GENDERHH" but there is no column in this data called "GenderHH".

```{r data2,echo=FALSE}
DT::datatable(BeanSurvey)
```

(You can use the arrow keys on your keyboard to scroll right in case the data table does not fit entirely on your screen)

